Fusing Filters with Integer Linear Programming

### 2014-04-22 2014betterfusionforfilters/section/03-SizeInference.tex
Refactor rules to generate constraints  Rename rate -> size consistently.  Use u^k as a size variable generator.  Make the main interpretation function take a list of bindings directly.  Add an interpretation for the top level function form.

%!TEX root = ../Main.tex
\section{Size inference}

Knowing the relative sizes of loops is important for fusion.
While it \emph{is} technically possible to fuse loops of different sizes, it requires extra complexity to find the maximum of the loop sizes, and additional branches are required to only execute the smaller loop fewer times. For simplicity, we do away with this added complexity and only support fusion of equal sized loops. \emph{Size inference} is performed on the combinators to infer as much information as possible about the sizes of the resulting loops.

@@

\subsection{Constraint generation}
Rates may either be some rate variable, $k_n$ where $n$ is the name of an array, or they may be a cross product of two rates.

\begin{tabbing}
MMMMM       \= MM \= MMMMMMMMMMM \= \kill
$rate$      \> @::=@ \> $k_n$               \> (rate variable)\\
            \> $~|$  \> $rate \times rate$  \> (cross product) \\
%            \> $~|$  \> $Filtered_n$        \> (output of @filter@s) \\
%            \> $~|$  \> $External_n$        \> (output of @external@s) \\
\end{tabbing}

The most important constraint is equality on two rates.
Context holes are used to simplify later definitions, and the notation $\fillhole{c}{d}$ is used to denote replacing all holes in $c$ with $d$.
Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be satisfiable.
Forall quantifiers may be given an upper bound, and mean the constraint must be satisfiable for every rate.
Existential quantifiers simply state that there must exist a rate that satisfies the constraint.

\begin{tabbing}
MMMMM       \= MM \= MMMMMMMMMMM \= \kill
$C$          \> @::=@ \> $true$                                 \> (trivially true) \\
             \> $~|$  \> $\hole$                                \> (constraint hole) \\
             \> $~|$  \> $rate = rate$                          \> (equality constraint) \\
             \> $~|$  \> $C \wedge C$                           \> (conjunction) \\
             \> $~|$  \> $\forall k_n.\ C$                      \> (forall quantification) \\
             \> $~|$  \> $\exists k_n.\ C$                      \> (existential quantification)\\
%             \> $~|$  \> $\exists k_n \le rate.\ C$             \> (exists with upper bound)    \\
\end{tabbing}

@@

$$\begin{array}{lcrcl}
&&\constr{\_} & :: & binds \rightarrow C \\
\\
% MMMMMMMMMMMMMMMMMM \= MM \= \kill
% $\constr{\lambda scalars~arrays \to @let @binds@ in @x}$ \> $=$ \> $\forall_{a \in arrays} k_a. \constr{binds}$ \\

\constr{o &=& @fold @f@ @n}       &  = &  \hole \\
\constr{o &=& @map@_n~f~ns}       &  = &  \exists k_o.\ \bigwedge_{n \in ns}\{k_o = k_n\} \wedge \hole \\
\constr{o &=& @filter@~f~n}       &  = &  \forall k_o.\ \hole \\
\constr{o &=& @gather@~i~d}       &  = &  \exists k_o.\ k_o = k_i \wedge \hole \\
\constr{o &=& @cross@~a~b}        &  = &  \exists k_o.\ k_o = k_a \times k_b \wedge \hole \\
\constr{outs &=& @external@~ins}  &  = &  \forall_{o \in outs} k_o.\ \hole \\
\\
&&\constr{b;~bs}  &  = &  \fillhole{\constr{b}}{\constr{bs}}       \\
&&\constr{nil}    &  = &  true                          \\
\end{array}$$

\subsubsection{Examples}

As an example of constraint generation, here is the @normalize2@ program from earlier.
\begin{tabbing}
@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
@normalize2 us@                     \> $\exists k_{us}.$      \\
@ = let sum1 = fold   (+) 0 us@     \>                      \\
@       gts  = filter (>0)  us@     \> $\forall k_{gts}.$ \\
@       sum2 = fold   (+) 0 gts@    \> \\
@       nor1 = map  (/sum1) us@     \> $\exists k_{nor1}.\ k_{nor1} = k_{us} \wedge$ \\
@       nor2 = map  (/sum2) us@     \> $\exists k_{nor2}.\ k_{nor2} = k_{us} \wedge$ \\
@   in (nor1, nor2)@                \> $true$ \\
\end{tabbing}
This constraint is satisfiable, with the equivalence classes being:
\newcommand{\eqclasses}[1]{
    \begin{tabbing}
        MM \= M \= \kill
        #1
    \end{tabbing}}
\newcommand{\eqclass}[2]{$#1$ \> $\in$ \> $\{#2\}$ \\}
\eqclasses{
    \eqclass{k_{us}}{k_{us}, k_{nor1}, k_{nor2}}
    \eqclass{k_{gts}}{k_{gts}}
}

>>>>

%!TEX root = ../Main.tex
\section{Size inference}
Knowing the relative sizes of loops is important for fusion.
While it \emph{is} technically possible to fuse loops of different sizes, it requires extra complexity to find the maximum of the loop sizes, and additional branches are required to only execute the smaller loop fewer times. For simplicity, we do away with this added complexity and only support fusion of equal sized loops. \emph{Size inference} is performed on the combinators to infer as much information as possible about the sizes of the resulting loops.

@@

% -----------------------------------------------------------------------------
\subsection{Constraint generation}
Sizes may either be some variable, $k_n$ where $n$ is the name of an array, or they may be a cross product of two rates.
\begin{tabbing}
MMMMM   \= MM \= MMMMMMMMM \= \kill
$size$  \> @::=@ \> $k_n$               \> (size variable) \\
        \> $~|$  \> $size \times size$  \> (cross product)
\end{tabbing}

The most important constraint is equality on two sizes.
Context holes are used to simplify later definitions, and the notation $\fillhole{c}{d}$ is used to denote replacing all holes in $c$ with $d$.
Constraints may be joined together with $c \wedge d$, which requires both $c$ and $d$ to be satisfiable.
Forall quantifiers may be given an upper bound, and mean the constraint must be satisfiable for every rate.
Existential quantifiers simply state that there must exist a rate that satisfies the constraint.
\begin{tabbing}
MMMMM   \= MM \= MMMMMMMMM \= \kill
$C$     \> @::=@ \> $true$              \> (trivially true)             \\
%       \> $~|$  \> $\hole$             \> (constraint hole)            \\
        \> $~|$  \> $size = size$       \> (equality constraint)        \\
        \> $~|$  \> $C \wedge C$        \> (conjunction)                \\
        \> $~|$  \> $\forall k_n.\ C$   \> (forall quantification)      \\
        \> $~|$  \> $\exists k_n.\ C$   \> (existential quantification)
\end{tabbing}

@@

$$\begin{array}{lcrcl}
&&\constr{\_} & :: & \textrm{function} \rightarrow C
\\[1ex]

\multicolumn{3}{l}{
    \constr{
       f~ \{ x_i \}~ \{ u_j \}^{\; j \gets 1..m}~ 
         =  \textrm{let}~ bs~ \textrm{in}~ exp
    }
}
        & = & \{ \exists u_j^k. \}^{j \gets 1..m}
           ~\constr{bs}

\\[2ex]

&&\constr{\_} & :: & \overline{bind} \rightarrow C 
\\[1ex]

&&\constr{~ \cdot ~}    
        & = &   true                          
\\[0.5ex]

\constr{u_o &=& @fold@~ f~ u_1;~ bs}
        & = &   \constr{bs}
\\[0.5ex]

\constr{u_o &=& @map@_n~ f~ \{u_i\}^{\;i \gets 1..n};~ bs }       
        & = &   \exists u_o^k.\ \bigwedge_{i \gets 1..n}\{u_o^k = u_i^k\} 
                \wedge \constr{bs}
\\[0.5ex]

\constr{u_o &=& @filter@~ f~ u_1;~ bs}
        & = &   \forall u_o^k.\ \constr{bs}
\\[0.5ex]

\constr{u_o &=& @gather@~ u_{ix}~ u_{data};~ bs}
        & = &   \exists u_o^k.\ u_o^k = u_{ix}^k
                ~\wedge~ \constr{bs}
\\[0.5ex]

\constr{u_o &=& @cross@~ u_a~ u_b;~ bs}
        & = &   \exists u_o^k.\ u_o^k = u_a^k \times u_b^k 
                ~\wedge~ \constr{bs}
\\[0.5ex]

\multicolumn{5}{l}{
\constr{x,~ \{u_i\}^{i \gets 1..n} = @external@~ \{v_j\}^{j \gets 1..m};~ bs}
          =     \{ \forall u_i^k. \}^{\; i \gets 1..n}
                ~\constr{bs}
}
\\[0.5ex]
\end{array}$$

% -----------------------------------------------------------------------------
\subsubsection{Examples}
As an example of constraint generation, here is the @normalize2@ program from earlier.
\begin{tabbing}
@MMMMMMMMMMMMMMMMMMMMMMMMMMMMMMMM@  \= \kill
@normalize2 us@                     \> $\exists k_{us}.$      \\
@ = let sum1 = fold   (+) 0 us@     \>                      \\
@       gts  = filter (>0)  us@     \> $\forall k_{gts}.$ \\
@       sum2 = fold   (+) 0 gts@    \> \\
@       nor1 = map  (/sum1) us@     \> $\exists k_{nor1}.\ k_{nor1} = k_{us} \wedge$ \\
@       nor2 = map  (/sum2) us@     \> $\exists k_{nor2}.\ k_{nor2} = k_{us} \wedge$ \\
@   in (nor1, nor2)@                \> $true$ \\
\end{tabbing}
This constraint is satisfiable, with the equivalence classes being:
\newcommand{\eqclasses}[1]{
    \begin{tabbing}
        MM \= M \= \kill
        #1
    \end{tabbing}}
\newcommand{\eqclass}[2]{$#1$ \> $\in$ \> $\{#2\}$ \\}
\eqclasses{
    \eqclass{k_{us}}{k_{us}, k_{nor1}, k_{nor2}}
    \eqclass{k_{gts}}{k_{gts}}
}
